\chapter{PREFACE TO THE THIRD RUSSIAN EDITION}
THE previous edition of this volume was the last book on which I worked together with my teacher L. D. Landau. The revision and expansion that we then carried out was very considerable, and affected every chapter.

For the third edition, naturally, much less revision was needed. Nevertheless, a fair amount of new material has been added, including some more problems, and relating both to recent research and to earlier results that have now become of greater significance.

Landau’s astonishing grasp of theoretical physics often enabled him to dispense with any consultation of original papers: he was able to derive results by methods of his own choice. This may have been the reason why our book did not contain certain necessary references to other authors. In the present edition, I have tried to supply them as far as possible. I have also added references to the work of Landau himself where we describe results or methods that are due to him personally and have not been published elsewhere.

As when dealing with the revision of other volumes in the Course of Theoretical Physics, I have had the assistance of numerous colleagues who informed me either of deficiencies in the treatment given previously, or of new material that should be added. Many useful suggestions incorporated in this book have come from A. M. Brodskii, G. F. Drukarev, I. G. Kaplan, V. P. Krainov, I. B. Levinson, P. E. Nemirovskii, V. L. Pokrovskii, I. I. Sobel’man, and I. S. Shapiro. My sincere thanks are due to all of these.

The whole of the work on revising this volume has been done in close collaboration with L. P. Pitaevskii. In him I have had the good fortune to find a fellow-worker who has passed likewise through the school of Landau and is inspired by the same ideals in the service of science.

\vspace{4ex}
\noindent E.M. LIFSHITZ, Moscow



\noindent\date{November 1973}